Prediction of Tuberculosis from Clinical Data

### 2024-04-07 code/eda-code/eda.qmd
part 4

figure_file = here("results","figures", "age_oxygenconsscatter_plot.png")
ggsave(filename = figure_file, plot=scatter_plot) 
```

# Create gtsummary table 
table1 <- 
  madaproject %>%
  tbl_summary(
    include = all_of(c("Gender", "Asthma", "Smoke", "Alcohol", "feverstatus", 
                       "Weightloss", "coughstatus", "Sputumproduction", 
                       "BloodinSputum", "cxrconclusion_chestxray", 
                       "homefuel", "hivstatus", "tb")),
    missing = "no",
    label = list(
      Asthma ~ "Asthma",
      Smoke ~ "Smoking Status",
      Alcohol ~ "Alcohol",
      feverstatus ~ "Fever Status",
      Weightloss ~ "Weight Loss",
      coughstatus ~ "Cough Status",
      `Sputumproduction` ~ "Sputum Production",
      `BloodinSputum` ~ "Blood in Sputum",
      cxrconclusion_chestxray ~ "CXR Conclusion",
      homefuel ~ "Home Fuel",
      hivstatus ~ "HIV Status",
      tb ~ "TB"
    )
  ) %>%
  add_n()

# Print the table
table1

# Save table
saveRDS(table1, file = here::here("results", "tables", "table1.rds"))

```
This is just my code to use a for loop 
```{r}

>>>>

figure_file = here("results","figures", "age_oxygenconsscatter_plot.png")
ggsave(filename = figure_file, plot=scatter_plot) 
```
Creating a table of characteristics

```{r}
library(dplyr)
library(gtsummary)

# Convert non-numeric variables to character
madaproject <- madaproject %>%
  mutate(across(where(is.factor), as.character))

# Define variables of interest (excluding "Age" and "oxygencons")
variables <- c("Gender", "Asthma", "Smoke", "Alcohol", "feverstatus", 
               "Weightloss", "coughstatus", "Sputumproduction", 
               "BloodinSputum", "cxrconclusion_chestxray", 
               "homefuel", "hivstatus")

# Create a summary table with specified variable types
summary_table_characteristics <- madaproject %>%
  select(all_of(variables), tb) %>%
  tbl_summary(
    by = tb,
    type = everything() ~ "categorical"
  )

# Print the summary table
summary_table_characteristics

# Save table
saveRDS(table1, file = here::here("results", "tables", "summary_table_characteristics .rds"))

This is just my code to use a for loop 
```{r}

### 2024-04-07 products/manuscript/Manuscript.qmd
part 4

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Summary/Abstract

@@

Tuberculosis (TB) remains a significant global health challenge due to diagnostic limitations affecting both pediatric and adult populations, resulting in delayed diagnosis or misdiagnosis[@Xin2019]. This delay impacts individual prognosis and community transmission. TB, caused by the bacterium Mycobacterium tuberculosis, primarily affects the lungs but can also involve other organs in its active form or remain latent without symptoms[@Bulled2020],. In 2020, TB ranked among the top 10 causes of death globally, with over 10 million new cases and 1.3 million deaths reported.

@@

What is the performance of the TB prediction models? {{< pagebreak >}}

@@

```{r}
table1 <- readRDS(here::here("results", "tables", "table1.rds"))
table_gt <- gtsummary::as_gt(table1)
table_gt |> gt::as_raw_html()
```

Figure1. Age distribution by TB

```{r}
#Figure1. Age distribution by TB
knitr::include_graphics(here("results","figures", "age_boxplot.png"))
```

figure2. TB and Age

## Basic statistical analysis

Model 1 which does not have cross validation has a very low ROC AUC value (12.21%), indicating poor discrimination between positive and negative instances. This suggests that Model 1's performance in distinguishing between the two classes is not much better than random guessing. Model 2 which uses cross validation, on the other hand, exhibits a significantly higher ROC AUC value (86.78%), suggesting much better performance in distinguishing between the positive and negative instances. A value of 86.78% indicates strong discriminatory power, with higher true positive rates and lower false positive rates across various thresholds. talk about the model number of folds

#Figure1. Age distribution by TB knitr::include_graphics(here("results","figures", "Roccurve.png"))

@@

>>>>

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(magrittr)
library(flextable)
library(htmltools)
```

# Abstract

#Introduction Although early detection and treatment of Tuberculosis cases are the hallmark of successful TB control, diagnostic delay is still long and common in Uganda. Therefore, the study aimed atdeveloping a tool that uses machine-learning techniques to detect and diagnose TB more accurately in a shorter time. #Methods This was a retrospective study that used secondary data collected between 2011 to 2018 collected from health under Infectious diseases research collaboration an (IDRC) project called MIND (Mulago Inpatient Non-invasive Diagnosis for pneumonia). The study was conducted on data from 2296 tested patients of whom 1345 had no TB and 951 had TB. Different machine learning metrics(Accuracy, RMSE and AUC) were used evaluate the performance of the models. #Results {{< pagebreak >}}

@@

Tuberculosis (TB) remains a significant global health challenge due to diagnostic limitations affecting both pediatric and adult populations, resulting in delayed diagnosis or misdiagnosis[@Xin2019]. This delay impacts individual prognosis and community transmission[@f_verbeke_4_2012]. TB, caused by the bacterium Mycobacterium tuberculosis, primarily affects the lungs but can also involve other organs in its active form or remain latent without symptoms[@Bulled2020],. In 2020, TB ranked among the top 10 causes of death globally, with over 10 million new cases and 1.3 million deaths reported.

@@

What is the performance of the TB prediction models?

@@

```{r}
#| label: tbl-characteristics
#| tbl-cap: " Participants' Characteristics"
#| echo: FALSE
ch<- readRDS(here("results", "tables", "characteristics.rds"))

ch %>%
  kable()
```

##Determining requirements for Tuberculosis prediction that is appropriate data and modeling techniques

In our study, we utilized 2296 instances and examined 14 variables associated with TB. Subsequently, nine variables were selected for prediction after conducting chi-square tests. Non-statistically significant variables associated with TB were eliminated. Following this, chi-square tests were applied to the nine chosen variables to assess their association with a threshold of 0.05. However, these tests did not reveal a significant association among the variables themselves. Therefore, they were included in the prediction model. The variables retained for prediction were: asthma, alcohol use, fever, chest radiography, home fuel use, weight loss, presence of blood in sputum, oxygen saturation, and HIV status.

```{r}
#| label: tbl-schematic2b
#| tbl-cap: "Model_perfomance."
#| echo: FALSE
mpf <- readRDS(here("results", "tables", "model_perfomance_train.rds"))

mpf %>%
  kable()
```

This study examines the efficacy of four distinct machine learning models in predicting the presence of tuberculosis (TB) using input data. Each model, including LASSO, Random Forest, Support Vector Machine (SVM), and a Null model, is rigorously evaluated based on key performance metrics. The LASSO model applies the Least Absolute Shrinkage and Selection Operator regularization technique, while the Random Forest model employs ensemble learning. SVM seeks to optimize hyperplanes for class separation, and the Null model serves as a simplistic baseline. Performance assessment encompasses accuracy (with LASSO achieving 80.26%, Random Forest achieving 81.57%, SVM achieving 79.68%, and the Null model achieving 60.23%), Area Under the Curve (AUC) (with LASSO achieving 0.8895, Random Forest achieving 0.8868, SVM achieving 0.8825, and the Null model achieving 0.5). The findings underscore the potential of advanced machine learning algorithms, particularly Random Forest and LASSO, in enhancing TB diagnosis and early detection strategies.

```{r}
#| label: tbl-Random Forest
#| tbl-cap: " Model Predictions (Random Forest)"
#| echo: FALSE
rf <- readRDS(here("results", "tables", "rfconf_table.rds"))

rf %>%
  kable()
```

The table presents the classification results of a Random Forest model for predicting tuberculosis (TB) status in individuals. Among individuals with TB (TB=YES), the model correctly identified 619 true positives and incorrectly predicted 219 cases as TB false negatives. For individuals without TB, the model accurately classified 709 cases as TB true negatives but misclassified 58 cases as TB false positives.

```{r}
#| label: fig-rf_ROCCURVE
#| fig-cap: "rf_ROCCURVE"
#| echo: FALSE
knitr::include_graphics(here("results","figures", "rf_ROCCURVE.png"))
```

```{r}
#| label: SVM
#| tbl-cap: "  Model Predictions (SVM)"
#| echo: FALSE
svm<- readRDS(here("results", "tables", "svmconf_table.rds"))
svm %>%
  kable()
```

The table presents the classification results of a Support Vector Machine model for predicting tuberculosis (TB) status in individuals. Among individuals with TB, the model correctly identified 621 true positives and incorrectly predicted 291 cases as TB false negatives. For individuals without TB, the model accurately classified 637 cases as TB true negatives but misclassified 56 cases as TB positive (false positives).

```{r}
#| label: fig-svm_ROCCURVE
#| fig-cap: "svm_ROCCURVE"
#| echo: FALSE
knitr::include_graphics(here("results","figures", "svm_ROCCURVE.png"))
```

```{r}
#| label: LASSO
#| tbl-cap: "  Model Predictions (LASSO)"
#| echo: FALSE
ls<- readRDS(here("results", "tables", "lassoconf_table.rds"))
ls %>%
  kable()
```

The table presents the classification results of a LASSSO model for predicting tuberculosis (TB) status in individuals. Among individuals with TB, the model correctly identified 654 true positives and incorrectly predicted 291 cases as TB false negatives. For individuals without TB, the model accurately classified 603 cases as TB true negatives but misclassified 23 cases as TB false positives.

```{r}
#| label: fig-Lasso_ROCCURVE
#| fig-cap: "Lasso_ROCCURVE"
#| echo: FALSE 
knitr::include_graphics(here("results","figures", "Lasso_ROCCURVE.png"))
```
```{r}
#Figure1. Random Forest ROC Curve

```

#Models Evaluation {{< pagebreak >}}

@@

The aim of the study was to apply machine-learning techniques in developing models for TB screening and detection in a shorter time based on patient data records to contribute to disease notification. The application of machine learning approaches have therefore been found to be successful in prediction [@Holzinger].
